---
title: "Ellerbe Creek Cleanup Tutorial"
author: "Margaret Swift, Jonathan Behrens"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  prettydoc::html_pretty:
    toc: true
    df_print: kable
---

# Welcome!

Welcome to the Ellerbe Creek Cleanup Tutorial! To download this file and follow along, head over to our [GitHub page](https://github.com/margaret-swift/ellerbe-creek)

# SECTION 01: WORKFLOW

### Set up your environment
Some basics: To run a block of code, click the green arrow in the top right of the gray box. 

Before we start, you need to set up your "environment." This involves installing and loading any libraries that you need to run the code for this module. Libraries basically store a bunch of really useful functions that other people have developed to make your life easier. You can both download and load packages with the `p_load` function! 

```{r load libraries}
# Load libraries (which have all the important functions). 
# The first line of code only needs to be ran once to load the package that 
# has the p_load function! Run it by removing the "#" before clicking the play button

#install.packages("pacman")
pacman::p_load(tidyverse, lubridate, dataRetrieval)

# Ensure the computer knows where to look for your files
setwd(dirname(rstudioapi::documentPath()))

```

### Load your data

Now that you've loaded all the libraries we need, the next step is to load all data for this project.

Notice that we are defining various variables. The following `a <- 3` means that we are telling R that we want the variable `a` to equal the number 3 (we call this "assigning a variable"). You can do this with numbers, text, and datasets, you name it!

```{r load data}
# This tells R to look for the folders and then giving them a name
data.dir <- "../01_data" # This folder has all the data
meta.dir <- "../00_meta" # This folder has all the meta data

# Do this for each dataset, to load it in
test<-file.path(data.dir, "durham_data_tutorial_version.csv") %>% # Finds file name
  read.csv() # Reads it into R
```

If you are new to R, maybe you have never seen the following before: %>%. This is called piping and is a powerful way to simplify your code and make it easier to read. Here's a wonderful [tutorial](https://medium.com/the-codehub/beginners-guide-to-piping-data-in-r-444054cbe999).

# SECTION 02: Why R?

### Reproducibility
By working in R, we have a record of every change or calculation we did on a dataset. This is very important so your work can be repeated by others. Also so you can repeat it if you find you accidentally made a mistake!

### Open Source = Accessible
R is open source. That means it is free and will likely always be! That means anyone can use your code and run it without paying anything for a software. This makes your work accessible to many.



# SECTION 03: Cleaning Up


### Open the durham_data and take a look

Now, back to the dataset at hand. Did you know that you can view data in R just like you do in Excel? Using the `View()` command, you can view the dataset like a spreadsheet. The `head()` function will show you just the top of the table (so it isn't too overwhelming). Try it out:

```{r view durham_data, eval=FALSE}

print(durham_data)

head(durham_data)

```

Wait a minute, something is not right. What's going on? Welcome to the fun of coding!

Sometimes, inappropriate metadata is stored within a .csv file. CSV, or comma-separated values, should only have data in them. If you open this spreadsheet in Excel, you will see that there is a set of data on top of the original that gives valuable metadata on the collection. We want to preserve this data, as it is important, but in a more appropriate way. An easy way to do this is to read the data in three parts. Once, to grab the metadata; second, to grab the column headers; third, to grab the actual data:

```{r fix metadurham_data}

# load the metadata (extra data at the top of the file) separately using "readLines"
meta <- readLines( file, n=10) # the metadata stops on row 10
meta <- paste( gsub(',', '', meta), collapse="\n") #remove errant commas
print(meta)

# now load the durham_data
durham_data <- read.csv(file, skip=12, row.names=NULL) #durham_data is on row 12+
head(durham_data)
```

Now we have the durham_data we need saved in 'durham_data', and the metadurham_data stored in 'meta'. Let's save the metadurham_data into a new file (NOT overwriting the raw durham_data).

```{r save metadurham_data}

meta.file.name <- 'metadurham_data.txt'
meta.file <- file.path(meta.dir, file.name)
write.table(meta, file=meta.file, row.names=FALSE, col.names=FALSE)

```


For data cleaning, we're going to cover three main issues: (1) Inconsistent Data Entry, (2) Duplicate Rows, and (3) Missing Data.


### INCONSISTENT DATA ENTRY

Now we are ready to look at the real data! It can be helpful at first to look at the "unique" data values for each column, to get a sense of what you're working with. Let's start with "Sky.Condition"

```{r explore durham_data with multiple spellings}
sky_condition<-durham_data %>% 
<<<<<<< HEAD
  group_by(Sky.Condition) %>% # group the data so all like things in "Sky.Condition" are put together
  count() %>% # counts the number of occurrences for a given type of "Sky Condition" 
  arrange(desc(n)) # arranges the rows by highest to lowest value of "n"
=======
  # group the data so all like things in "Sky.Condition" are put together
  group_by(Sky.Condition) %>% 
  # count the number of occurances for a given type of "Sky Condition" 
  count() %>% 
  # arrange the rows by highest to lowest value of "n"
  arrange(desc(n)) 
>>>>>>> f8df9e8960052e79a71b3463023d9fc324c5c7df

print(sky_condition) # shows the result!

```

Ah, it looks like somebody had some inconsistencies with spellings. 

```{r}

# First, make everything in the column lowercase 'tolower()' to make your life easier

durham_data <- durham_data %>%
  mutate(Sky.Condition = tolower( Sky.Condition ))

sort( unique( durham_data$Sky.Condition ) )

durham_data <- durham_data %>%
  mutate(Sky.Condition = tolower( Sky.Condition ))

```



### DUPLICATE durham_data
 
```{r find duplicates}

# find rows with "duplicate" in the comments, then create a new durham_data frame just with those.
inx.dup <- which( grepl( 'duplicate', tolower(durham_data$Comments) ) ) 
dupes <- durham_data[inx.dup,] 
head( dupes )

```

Normally, you could just get rid of duplicates using the function distinct(). distinct() simply drops the second row that is duplicated, however. Take a look at some of these values in 'dupes' -- they're not the same! When looking at your own data, you need to make decisions on how to deal with duplicate data. For now, we've decided to take the average of the two values. If you want to use just the first value, distinct() is fine; otherwise, here's how to replace duplicate values with the averages:

```{r}

# Create an ID row to sort on station name, filtered, parameter, and date
dupes <- dupes %>% 
  mutate(ID = paste0(Station.Name, Filtered, Parameter, Date.Time)) %>%
  arrange(ID)

# grab the mean values for each combination
means <- dupes %>% 
  group_by(ID) %>% #group dataframe by ID
  summarize(MeanValue = mean(Value, na.rm=TRUE)) #take the mean and ignore NAs

# collapse the duplicate data frame based on ID and sort it the same way
dupes <- dupes %>% 
  distinct(Station.Name, Filtered, Parameter, Date.Time, .keep_all=TRUE)

# make sure both datasets are the same length and in the same order
nrow(dupes) == nrow(means) && all(dupes$ID == means$ID)

# Now that we're sure, replace Value column of distinct dupes with mean values 
# and remove the ID column
dupes$Value <- means$MeanValue
dupes <- dupes %>% dplyr::select(-ID)

# Put it all together! First, remove duplicated rows from the main data frame
durham_data <- durham_data[-inx.dup,]

# Add the averaged, previously-duplicated rows to the end of the data frame
durham_data <- bind_rows(durham_data, dupes) %>%
  arrange(Station.Name, Filtered, Parameter, Date.Time) #sort data again

head(durham_data)
```



### MISSING DATA

```{r}
summary( durham_data$Value )
any( is.na(durham_data$Value) )

```


### DATA FORMAT

Now it's time to get the data ready for visualizations. 

```{r}
# Let's select just a few columns of data using the "select" function. 
D1 <- durham_data %>%
  dplyr::select(Station.Name, Date.Time, Parameter, Value) 

# Now let's filter for distinct rows. We can do this first by grouping by Name and Date. 
# Put simply, R will basically take all the rows from the same date, then look at rows with the same Station.Name, and finally look for duplicates and remove. 
D2<-D1 %>%
  group_by(Station.Name, Date.Time) %>%
  distinct(Station.Name, Date.Time, Parameter, .keep_all=TRUE) 

# Next we are going to make the dataset "wider". Transforming the dataset.
D3<-D2%>%
  pivot_wider(names_from=Parameter, values_from=Value) 

# And we will finish by cleaning it up
durham_data_wide<-D3 %>% 
  mutate(Date = as.Date(Date.Time, format="%m/%d/%y"),
         Year = year(Date),
         Date = as.POSIXct.Date(Date)) %>%
  relocate(Date, Year, .after=Date.Time) %>%
  ungroup()

# Remove D1, D2, and D3 since durham_data_wide is all we really want 
rm(D1, D2, D3)

```


# SECTION 04: Visualization and Analysis

Now let's visualize some of this data! Through visualizations, we can take complex or large datasets and simplify them to look for trends. That in turn can help guide our analysis or direct us towards interesting and new questions. 

### Time Series Plots and Analysis
Starting with the basics, we can think about this dataset as a time series and plot a given analyte against time. 

What analytes exist? Well they are the names of the columns. Let's look at their names
```{r}
colnames(durham_data_wide)
```


Let's see how concentrations of calcium changes over time. We can color the points based on the site where the sample was taken.

```{r}

ggplot() +
  geom_point(data=durham_data_wide, mapping=aes(x=Date, y=Calcium, color=Station.Name)) +
  theme_classic()


```

Using similar code to above, choose two different variables and see how they compare to one another! 

Hint: You can simply change the x-axis from "Date" to your variable of choosing.
```{r}
# INSERT YOUR CODE HERE


```



Cool, we can see how concentration varies overtime, but it's hard to tell how variation throughout the year differs between sites. So let's look at a boxplot of this data!

```{r}

ggplot(
  data=durham_data_wide, aes(x=Station.Name, y=Calcium)) +
  geom_boxplot()
  
```

Let's get fancy and separate this out by year. Maybe there's a difference between years?

```{r}

ggplot(
  data=durham_data_wide, aes(x=Station.Name, y=Calcium)) +
  geom_boxplot() +
  facet_wrap(~Year)
  
```

Ah shoot the graph is a little messy, let's clean it up
```{r}

ggplot(
  data=durham_data_wide, aes(x=Station.Name, y=Calcium)) +
  geom_boxplot() +
  facet_wrap(~Year) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
  
```

Try doing the same visualizations now, but with a different variable. Make sure you use the exact name of the variable as it is found in the dataset!

```{r}
# INSERT YOUR CODE HERE!



```


### Exploring a Federal Dataset

It just so happens that the City of Durham isn't the only one sampling Ellerbe Creek! We also have the US Geological Survey (USGS), which has a bunch of sensors deployed in streams and rivers nationally, and two in Ellerbe Creek. 

Let's explore what they have, using USGS' R package (dataRetrieval) to download and view the data.

First, let's introduce you to the wonderful world of CRAN. Every package will usually create a helpful documentation outlining every function in the package. It will include a list of every function, and a help page for each (usually with examples). The dataRetrieval documentation can be [found here](https://cran.r-project.org/web/packages/dataRetrieval/dataRetrieval.pdf).

``` {r}
# Look for all sites in NC that have measures for mean daily Discharge, cubic feet per second (code = 00060)
siteListDischarge <- whatNWISsites(stateCd="NC",parameterCd="00061") %>% 
  filter(str_detect(station_nm, "ELLERBE")) # Filter for sites that are for Ellerbe Creek

head(siteListDischarge)
```

Sweet! Let's grab the site near Club Boulevard and Gorman. The first is upstream of the second. Based on this, which would you expect to have a higher flow? Let's see if you are right!

First, load in the data
```{r}
siteNos <- c("0208675010", "02086849")
pCode <- "00060"

Discharge_Data <- readNWISdv(siteNos, pCode, "2005-01-01","2021-12-31")

```


Now let's take a look!
```{r}

ggplot() +
  geom_point(data=Discharge_Data, mapping=aes(x=Date, y=X_00060_00003, color=site_no)) +
  theme_classic() +
  labs(y="Average Daily Discharge (ft^3/sec)")


```

Yikes that is messy. Maybe we should zoom into one year, and then replace the dots with lines
```{r}
# Filter data
Discharge_Data_2020<-Discharge_Data %>% 
  filter(Date > "2020-01-01" & Date < "2020-12-31")

# Plot data
ggplot() +
  geom_line(data=Discharge_Data_2020, mapping=aes(x=Date, y=X_00060_00003, color=site_no)) +
  theme_classic() +
  labs(y="Average Daily Discharge (ft^3/sec)")

```



# SECTION 05: Conclusion
